Left-linear rules in adhesive categories

### 2024-02-07 icalp/fusione.tex
structure

  The converse holds only when rules are linear [proof clear, but not written]

\end{itemize}

@@

\todo{A VERY NICE INTRODUCTION}

SOME IDEAS (BADLY WRITTEN, JUST TO START TO PUT IDEAS IN ORDER)

Rewriting systems over graphs are used in uncountably many setting for
representing the dynamics of software systems: the graph serves to represent
the entities and their relations within states , while rewriting steps,
modifying the graph, model computational steps that results in
changes to the system state.

Rewriting over many kind of graphs is elegantly captured by the
so-called \emph{double pushout} (DPO) approach to rewriting. A
noteworthy advantage is that DPO rewriting can be formulated over
adhesive categories and their
variants~\cite{lack2005adhesive,ehrig2006weak}. Operating within the
framework of adhesive categories allows one to devise the foundations
of a rewriting theory which can be subsequently instantiated to the
context of interest, thus avoiding the need of repeatedly proving
similar ad-hoc results in for each specific setting.

Rewriting over adhesive categories has been widely studied in the last
two decades and most of the research focused on the so-called linear
rewriting rules. Roughly speaking, linear rewriting rules can specify
that, in a rewriting step, some entities in the current state are
removed, some other are needed for the rewriting step to occur but
remain unaltered and some new entities can be generated.

The computation described by a rewriting system can be naturally
interpreted as a distributed computation and a notion of independence
between rewriting steps naturally emerges. When different rules are
applied at ``distinct'' parts of the state, possibly overlapping on
preserved items, the rewriting steps can be naturally deemed as
independent and their application can occur in any order and possibly
concurrently. This phenomenon is formally captured by the so-called
Church-Rosser theorem of DPO rewriting~\cite{CMREHL:AAGT}.
%
Having a notion of independence between computation steps sits at the
core of the possibility of viewing an execution trace as a concurrent
computation. This is obtained by considering traces up to an
equivalence that identifies traces differing only for the order of
independent steps, referred to as shift equivalence~\cite{CMREHL:AAGT}
(similar to interchange law~\cite{Mes92} in term rewriting or
permutation equivalence~\cite{JJL80} in the $\lambda$-calculus).

In some situations rewriting with linear rules are not sufficiently
expressive for capturing the dynamic of a system. This is the case
when a computational step is required not only to consume, preserve
and generated, but also ``merge'' parts of the state. This happens,
e.g., in nominal calculi where, as a result of name passing, the
received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimizations~\cite{WNW:egg}. Rather than
explicitly merging nodes corresponding to equal expression, the idea
here - conceptually and, as we will see, also mathematically similar -
is to maintain an equivalence\todo{P: more precisely a congruence}
over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
  vedere se c'\`e qualcosa d'altro}

\todo{P: something about string diagrams?}

@@

\begin{notation} 
We stipulate here some notational conventions which will be used throughout this paper. 
\begin{itemize}\item 
	Given a category $\X$ we will not distinguish notationally between $\X$ and its class of objects: so that ``$X\in \X$'' means that $X$ belongs to the class of objects of $\X$.  
	\item 
	If $1$ is a terminal object in a category $\X$,  the unique arrow $X\to 1$ from another object $X$ will be denoted by $!_X$. Similarly, if $0$ is initial in $\X$ then $?_X$ will denote the unique arrow $0\to X$. %When $\X$ is $\Set$ and $1$ is a singleton, $\delta_x$ will denote the arrow $1\to X$ with value $x\in X$.
	\item  $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ will denote the class of all arrows, monos and regular monos of $\X$, respectively.

@@

\begin{example}\label{rem:iso} Let $\mathsf{Iso}$ be the class of all isomorphisms, then any category $\X$ is $\mathsf{Iso}$-adhesive.
\end{example}

@@

\begin{corollary}\label{cor:slice}
	Let  $X$ be an object of an $\mathcal{M}$-adhesive category $\X$. Then  $\X/X$ and $X/\X$ are, respectively, is $\mathcal{M}/X$- and $X/\mathcal{M}$-adhesive, where
	\[\mathcal{M}/X:=\{m\in  \mathcal{A}(\X/X) \mid m\in \mathcal{M} \} \qquad X/\mathcal{M}:=\{m\in  \mathcal{A}(X/\X) \mid m\in \mathcal{M} \}\]
\end{corollary}

@@

\begin{theorem}[\cite{ehrig2006fundamentals,lack2005adhesive}]\label{thm:functors}
If $\X$ is an $\mathcal{M}$-adhesive category, then for every small category $\Y$, the category $\X^\Y$  of functors $\Y\to \X$ is $\mathcal{M}^{\Y}$-adhesive, where
\[\mathcal{M}^{\Y}:=\{\eta \in \mathcal{A}(\X^\Y) \mid \eta_Y \in \mathcal{M} \text{ for every } Y\in \Y\}\]
\end{theorem}

@@

\begin{example}\todo{GRAFI GERARCHICI}
\end{example}

\begin{example}\todo{term graph}
\end{example}

@@

	\noindent 
\parbox{10cm}{\begin{notation}
  Let $\der{D}=\{\dder{D}_i\}_{i=0}^n$ be a derivation. We will depict
  the $i^\text{th}$ element $\dder{D}_i$ of $\der{D}$ as in the
  diagram on the right. Notice that, in particular, if $\der{D}$ is a derivation between $G$ and $H$, then $G_0=G$ and $G_{n+1}=H$. When $\der{D}$ has length $1$ we will suppress the
  indexes. In such case, we will also identify $\der{D}$ with its only
  element.
\end{notation} } \parbox{3cm}{$\xymatrix{L_i \ar[d]_{m_i}& K_i \ar[d]^{k_i}\ar@{>->}[l]_{l_i} \ar[r]^{r_i} & R_i \ar[d]^{h_i} \\G_{i} & \ar@{>->}[l]^{f_{i}} D_{i} \ar[r]_{g_{i}}& G_{i+1} }$}

@@

\noindent 
\parbox{8cm}{
\begin{definition}Let $(\X, \R)$ be a left-linear DPO rewriting system,  an \emph{abstraction equivalence} between two derivations $\der{D}$ and $\der{D'}$ with the same length and such that $r(\der{D})=r(\der{D}')$, is a family of isomorphisms $\{\phi_X\}_{X\in \Deltamin(\der{D})}$ such that, for every $i\in [0, \lgh(\der{D})]$ the diagram on the right commutes. We will say that  $\der{D}$are $\der{D}'$  are \emph{abstraction equivalent}, if there exists an abstraction equivalence between them.
\end{definition}}
\parbox{4cm}{\vspace{-1.35em}$\xymatrix@C=40pt{G'_i&D'_i \ar[r]^{g'_i} \ar@{>->}[l]_{f'_i}&G'_{i+1}\\  L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i \ar[u]^{k'_i} \ar[d]_{k_i} \ar[r]^{r_i} \ar@{>->}[l]_{l_i} &R\ar[u]^{h'_i} \ar[d]_{h_i}\\G_i \ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole&D_i\ar@{>->}[l]^{f_i}\ar@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole \ar[r]_{g_i}&G_{i+1}\ar@/_.45cm/[uu]_{\phi_{H_i}}}$}

@@

\begin{theoremEnd}[category=sec2]{proposition}\label{prop:unique} Let $\X$  be a $\mathcal{M}$-adhesive category. Suppose that the two direct derivations $\dder{D}$ and $\dder{D}'$ below, with the same match and applying the same left-linear rule $\rho$ are given.
	\[\xymatrix{L \ar[d]_{m}& K \ar[d]^{k}\ar@{>->}[l]_{l} \ar[r]^{r} & R \ar[d]^{h} & L \ar[d]_{m}& K \ar[d]^{k'}\ar@{>->}[l]_{l} \ar[r]^{r} & R \ar[d]^{h'}\\G & \ar@{>->}[l]^{f} D \ar[r]_{g}& H & G & \ar@{>->}[l]^{f'} D' \ar[r]_{g'}& H'}\]
	Then there an abstraction equivalence between $\dder{D}$ and $\dder{D}'$, whose first component is $\id{G}$.
\end{theoremEnd}
\begin{proof}\begin{proofEnd}
		Both pairs $(k, f)$ and $(k', f')$ are pushout complements of $l$ and $n$, thus, \Cref{lem:pocomp} yields an isomorphism $\phi_D\colon D\to D'$. Computing we have
		\[
		g'\circ \phi_D \circ k= g' \circ k'=h'\circ r\]

		\noindent
		\parbox{3cm}{
			$\xymatrix{K  \ar@/^.4cm/[rr]^{k'}\ar[d]_{r} \ar[r]_{k}& \ar[r]_{\phi_D} D \ar[d]^{g}& D' \ar[d]^{g'}\\ R \ar@/_.4cm/[rr]_{h'} \ar[r]^{h}& H \ar[r]^{\phi_H}& H'}$} \hfill 
		\parbox{10cm}{
		\hspace{15pt}Hence, we have another $\phi_H\colon H\to H'$. To see that $\phi_H$ is an isomorphism, consider the diagram aside. By hypothesis the whole rectangle and its left half are pushouts, therefore, by \Cref{lem:po1} its right square is a pushout too. The claim now follows from the fact that the pushout of an isomorphism is an isomorphism. \qedhere }
	\end{proofEnd}
\end{proof}

\iffalse

\subsection{Consistent permutations}

@@

			Let $(C, \{c_X\}_{X\in \Delta(\der{D})})$ be a cocone, since $\Delta(\der{D}')$ is a subdiagram of $\Delta(\der{D})$, we get another cocone $(c, \{c_X\}_{X\in \Delta(\der{D}')})$ which induces an arrow $c'\colon \lpro \der{D}' \rpro \to C$ such that
			\begin{align*}
				c'\circ \iota_{G_n} \circ f_n &=c_{G_n} \circ f_n\\&= c_{D_n}\\&= c_{H}\circ g_n
			\end{align*}
			Therefore the arrows $c'$ and $c_H$ induce a morphism $c\colon P\to C$ and the thesis now follows at once.

@@

	Let us now show that the square above is a pushout. Take two arrows $a\colon \tproi{D}{1}\to C$, $b\colon \tproi{D}{2}\to C$ such that
	\[a\circ \iota_{1, H}=b\circ \iota_{2, H}\]
	We can use the previous equality to define a cocone $(C, \{c_X\}_{X\in \Delta(\der{D})})$ putting:
	\[c_X:=\begin{cases}
		a\circ \iota_{1, X} & X\in \Delta(\der{D}_1)\\
		b\circ \iota_{2, X} & X\in \Delta(\der{D}_2)
	\end{cases}\]
From this, we can deduce at once the existence of a unique $c\colon \tpro{D}\to C$ such that 
\[c\circ \iota_X = c_X\]
By construction, for every $X\in  \Delta(\der{D}_1)$ and $Y\in  \Delta(\der{D}_2)$ we have
\[\begin{split}
	c\circ p_1 \circ \iota_{1,X}&=c\circ \iota_{X}\\&=c_X \\&=a\circ \iota_{1,X}\\&=a\circ p_1\circ \iota_{1,X}
\end{split}\qquad \begin{split}
	c\circ p_2 \circ \iota_{2,Y}&=c\circ \iota_{Y}\\&=c_Y \\&=b\circ \iota_{2,Y}\\&=b\circ p_2\circ \iota_{2,Y}
\end{split}\]
Therefore
\[c\circ p_1=a \qquad c\circ p_2 = b\]

	For uniqueness, suppose that $c'\colon \tpro{D}\to C$ is such that
	\[c'\circ p_1=a \qquad c'\circ p_2 = b\]
Then, for every $X\in \Delta(\der{D})$ we have
\begin{align*}
	c'\circ \iota_X &= \begin{cases}
	c'\circ p_1\circ \iota_{1, X} & X\in \Delta(\der{D}_1)\\
	c'\circ p_2\circ \iota_{2, X} & X\in \Delta(\der{D}_2)
	\end{cases}\\&=\begin{cases}
a\circ \iota_{1, X} & X\in \Delta(\der{D}_1)\\
b\circ \iota_{2, X} & X\in \Delta(\der{D}_2)
	\end{cases}\\&=c_X\\&=c\circ \iota_X
\end{align*}
showing that $c'=c$ as wanted.	 \qedhere 
	\end{enumerate}
\end{proof}

\begin{corollary}\label{cor:colim}
Let $\der{D}=\{\dder{D}_i\}_{i=0}^n$ a derivation of length $n+1$ and fix an index $j\in[0,n]$. Define
\[\der{D}^j_1:=\{\dder{D}_i\}_{i=0}^{j-1} \qquad  \der{D}^j_2=\{\dder{D}_j\} \qquad \der{D}^j_3:=\{\dder{D}_i\}_{i=j+1}^n\]
with the convention that $\der{D}^0_1$ and $\der{D}^n_3$ are the empty derivation on, respectively, $G_0$ and $G_n$. Then the square below is a pushout and a pullback
\[\xymatrix@C=30pt{D_{j} \ar[ddrr]^{\iota_{D_j}} \ar[r]^{g_j}\ar@{>->}[d]_{f_j}& G_{j+1} \ar[ddr]^{\iota_{G_{j+1}}} \ar[r]^-{\iota_{3, G_{j+1}}} & \lpro \der{D}^j_3\rpro \ar[dd]^{p_2} \\ G_j\ar[drr]_{\iota_{G_{j}}} \ar@{>->}[d]_{\iota_{1,G_j}}\\ \lpro \der{D}^j_1 \rpro \ar[rr]_{p_1}  &&\tpro{D} }\] 
Where the two arrows $p_1\colon \lpro\der{D}^j_1 \rpro\to \tpro{D}$, $p_2\colon \lpro \der{D}^j_3\rpro \to \tpro{D}$ are induced by the cocones $(\tpro{D}, \{\iota_{X}\}_{X\in \Delta(\der{D}^j_1)})$ and $(\tpro{D}, \{\iota_{X}\}_{X\in \Delta(\der{D}^j_3)})$, respectively.
\end{corollary}
\begin{remark}
If $\der{D}$ is empty then $\der{D}^j_1, \der{D}^j_2$ and $\der{D}^j_3$ are empty too.
\end{remark}
\begin{proof}
We can notice that $\der{D}=\der{D}^j_1\cdot \der{D}^j_2 \cdot \der{D}^j_3$. By the first and the second point of \Cref{lem:colim} then we get the following diagram, in which all squares are $\mathcal{M}$-pushouts.

\[\xymatrix@C=30pt{D_{j}  \ar[r]^{g_j}\ar@{>->}[d]_{f_j}& G_{j+1}  \ar[r]^-{\iota_{3,G_{j+1}}} \ar[d]_{\iota_{2, G_{j+1}}}  \ar@/^.5cm/[dd]^{\iota_{1,2, G_{j+1}}}& \lpro \der{D}^j_3\rpro \ar[dd]^{p_2} \\ G_j \ar[r]^{\iota_{2, G_j}}\ar@{>->}[d]_{\iota_{1,G_j}} & \lpro \der{D}^j_2\rpro \ar[d]_a\\ \lpro \der{D}^j_1 \rpro \ar@/_.4cm/[rr]_{p_1} \ar[r]^b &\lpro \der{D}^j_1\cdot \der{D}^j_2 \rpro \ar[r]^c&\tpro{D} }\] 
Applying \Cref{lem:po1} twice we get that the whole square is an $\mathcal{M}$-pushout. Then the thesis follows from \Cref{prop:pbpoad}.
\end{proof}

\begin{remark}\label{rem:zero1} In particular, considering $j=0$ or $j=n$, we get that the following two squares are $\mathcal{M}$-pushouts and, thus, pullbacks.
	\[\xymatrix{D_0 \ar[r]^-{\iota_{3, D_0}} \ar@{>->}[d]_{f_0}& \lpro \der{D}^0_3 \rpro \ar@{>->}[d]^{p_2} &D_n \ar@{>->}[d]_{\iota_{1, D_n}}\ar[r]^{g_n}& G_n \ar@{>->}[d]^{\iota_{G_n}}\\ G \ar[r]_{\iota_G}& \tpro{D} & \lpro \der{D}^n_1 \rpro \ar[r]_{p_1}  & \tpro{D}}\]
\end{remark}

>>>>

  The converse holds only when rules are linear [proof clear, but not
  written]  
\end{itemize}

@@

%\todo{A VERY NICE INTRODUCTION}
%
%SOME IDEAS (BADLY WRITTEN, JUST TO START TO PUT IDEAS IN ORDER)

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
%performed by the device 
via a suitable equivalence. Intuitively, this equivalence 
captures when steps are causally unrelated and thus could be executed in any 
order, and possibly concurrently. 
%
Two seminal contributions to this line of research have been
Mazurkiewicz's traces~\cite{Mazurkiewicz86} and Winskel's event structures~\cite{NPW:PNES}
%
Concerning formalisms based on the rewriting paradigm, i.e. where the
behaviour is dictated by a set of rules which have to be instantiated with
the system at hand, concurrency often boils down to having a notion of 
independence between two consecutive steps.
%thus viewing a sequence of steps as a concurrent computation. 
As noteworthy examples, we have 
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.

Among rewriting-based formalisms, those manipulating graph-like structures
proved useful in many settings for representing the dynamics of systems: the 
graph serves to represent the entities and their relations within states, while 
rewriting steps, modifying the graph, model computation steps that results 
in changes to the system state. The DPO (double-pushout) approach~\cite{CMREHL:AAGT} 
is nowadays considered the standard one for these structures, thanks to its locality 
(rule application has a local effect on a state, differently e.g. from the SqPO 
approach~\cite{CorradiniHHK06}) and most importantly to its flexibility: the rules allow to 
specify that,  in a rewriting step, some entities in the current state are removed, 
some others are needed for the rewriting step to occur but remain unaltered and 
some new entities can be generated. Concerning concurrency, the chosen notion 
is shift equivalence, and independence is formally captured by the so-called 
Church-Rosser theorem of DPO rewriting~\cite[xxx]{CMREHL:AAGT}.

A noteworthy advantage is that DPO rewriting can be formulated over
adhesive categories and their variants~\cite{lack2005adhesive,ehrig2006weak}. 
%
Operating within the framework of adhesive categories allows one to devise 
the foundations of a rewriting theory which can be subsequently instantiated 
to the context of interest, thus avoiding the need of repeatedly proving
similar ad-hoc results in for each specific setting. In fact, the framework is
so general that it allowed to recast in the DPO approach many additional formalisms,
ranging from xxx to string diagrams~\cite{bonchi2022string}.

So far, most of the theoretical results focussed on what are called \emph{linear rules}: 
intuitively, a step is required only to consume, preserve and generate entities. 
However, in some situations it is also necessary to ``merge'' parts of the state. 
This happens already in the context of string diagram mentioned above, but also 
in graphical implementations of nominal calculi where, as a result of name passing, 
the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the visual modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expression, the idea
here - conceptually and, as we will see, also mathematically similar -
is to maintain an equivalence\todo{P: more precisely a congruence}
over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
  vedere se c'\`e qualcosa d'altro}

% =========DA QUI SUPERSEDED\\
% Rewriting systems over graphs are used in uncountably many setting for
% representing the dynamics of software systems: the graph serves to represent
% the entities and their relations within states , while rewriting steps,
% modifying the graph, model computational steps that results in
% changes to the system state.

% Rewriting over many kind of graphs is elegantly captured by the
% so-called \emph{double pushout} (DPO) approach to rewriting. A
% noteworthy advantage is that DPO rewriting can be formulated over
% adhesive categories and their
% variants~\cite{lack2005adhesive,ehrig2006weak}. Operating within the
% framework of adhesive categories allows one to devise the foundations
% of a rewriting theory which can be subsequently instantiated to the
% context of interest, thus avoiding the need of repeatedly proving
% similar ad-hoc results in for each specific setting.

% Rewriting over adhesive categories has been widely studied in the last
% two decades and most of the research focused on the so-called linear
% rewriting rules. Roughly speaking, linear rewriting rules can specify
% that, in a rewriting step, some entities in the current state are
% removed, some other are needed for the rewriting step to occur but
% remain unaltered and some new entities can be generated.

% The computation described by a rewriting system can be naturally
% interpreted as a distributed computation and a notion of independence
% between rewriting steps naturally emerges. When different rules are
% applied at ``distinct'' parts of the state, possibly overlapping on
% preserved items, the rewriting steps can be naturally deemed as
% independent and their application can occur in any order and possibly
% concurrently. This phenomenon is formally captured by the so-called
% Church-Rosser theorem of DPO rewriting~\cite{CMREHL:AAGT}.
% %
% Having a notion of independence between computation steps sits at the
% core of the possibility of viewing an execution trace as a concurrent
% computation. This is obtained by considering traces up to an
% equivalence that identifies traces differing only for the order of
% independent steps, referred to as shift equivalence~\cite{CMREHL:AAGT}
% (similar to interchange law~\cite{Mes92} in term rewriting or
% permutation equivalence~\cite{JJL80} in the $\lambda$-calculus).

% In some situations rewriting with linear rules are not sufficiently
% expressive for capturing the dynamic of a system. This is the case
% when a computational step is required not only to consume, preserve
% and generated, but also ``merge'' parts of the state. This happens,
% e.g., in nominal calculi where, as a result of name passing, the
% received name is identified with a local one at the
% receiver~\cite{CVY:ESSPE,Gad07} or in the modelling of bonding in
% biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
% at the core of e-graphs (equality graphs), recently used for
% rewrite-driven compiler optimisations~\cite{WNW:egg}. Rather than
% explicitly merging nodes corresponding to equal expression, the idea
% here - conceptually and, as we will see, also mathematically similar -
% is to maintain an equivalence\todo{P: more precisely a congruence}
% over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
%   vedere se c'\`e qualcosa d'altro}\\
% =========

@@

\begin{notation}
  We stipulate here some notational conventions which will be used
  throughout this paper.
  \begin{itemize}
  \item
    Given a category $\X$ we will not distinguish notationally
    between $\X$ and its class of objects: so that ``$X\in \X$'' means
    that $X$ belongs to the class of objects of $\X$.

  \item
    If $1$ is a terminal object in a category $\X$, the unique
    arrow $X\to 1$ from another object $X$ will be denoted by
    $!_X$. Similarly, if $0$ is initial in $\X$ then $?_X$ will denote
    the unique arrow $0\to
    X$. %When $\X$ is $\Set$ and $1$ is a singleton, $\delta_x$ will denote the arrow $1\to X$ with value $x\in X$.

  \item $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ will denote the class of
    all arrows, monos and regular monos of $\X$, respectively.

@@

\begin{example}
  \label{rem:iso}
  Let $\mathsf{Iso}$ be the class of all
  isomorphisms, then any category $\X$ is $\mathsf{Iso}$-adhesive.
\end{example}

@@

\begin{corollary}\label{cor:slice}
  Let $X$ be an object of an $\mathcal{M}$-adhesive category
  $\X$. Then $\X/X$ and $X/\X$ are, respectively, is $\mathcal{M}/X$-
  and $X/\mathcal{M}$-adhesive, where
  \[\mathcal{M}/X:=\{m\in \mathcal{A}(\X/X) \mid m\in \mathcal{M} \}
    \qquad X/\mathcal{M}:=\{m\in \mathcal{A}(X/\X) \mid m\in
    \mathcal{M} \}
  \]
\end{corollary}

@@

\begin{theorem}[\cite{ehrig2006fundamentals,lack2005adhesive}]
  \label{thm:functors}
  If $\X$ is an $\mathcal{M}$-adhesive category, then for every small
  category $\Y$, the category $\X^\Y$ of functors $\Y\to \X$ is
  $\mathcal{M}^{\Y}$-adhesive, where
  \[\mathcal{M}^{\Y}:=\{\eta \in \mathcal{A}(\X^\Y) \mid \eta_Y \in
    \mathcal{M} \text{ for every } Y\in \Y\}\]
\end{theorem}

@@

\begin{example}
  \todo{GRAFI GERARCHICI}
\end{example}

\begin{example}
  \todo{term graph}
\end{example}

@@

\noindent
\parbox{10cm}{%
  \begin{notation}
    Let $\der{D}=\{\dder{D}_i\}_{i=0}^n$ be a derivation. We will
    depict the $i^\text{th}$ element $\dder{D}_i$ of $\der{D}$ as in
    the diagram on the right. Notice that, in particular, if $\der{D}$
    is a derivation between $G$ and $H$, then $G_0=G$ and
    $G_{n+1}=H$. When $\der{D}$ has length $1$ we will suppress the
    indexes. In such case, we will also identify $\der{D}$ with its
    only element.
  \end{notation} }
\parbox{3cm}{
  $\xymatrix{L_i \ar[d]_{m_i}& K_i
    \ar[d]^{k_i}\ar@{>->}[l]_{l_i} \ar[r]^{r_i} & R_i \ar[d]^{h_i}\\
    G_{i} & \ar@{>->}[l]^{f_{i}} D_{i} \ar[r]_{g_{i}}& G_{i+1} }$
}

@@

\noindent
\parbox{8cm}{
  \begin{definition}
    Let $(\X, \R)$ be a left-linear DPO rewriting system, an
    \emph{abstraction equivalence} between two derivations $\der{D}$
    and $\der{D'}$ with the same length and such that
    $r(\der{D})=r(\der{D}')$, is a family of isomorphisms
    $\{\phi_X\}_{X\in \Deltamin(\der{D})}$ such that, for every
    $i\in [0, \lgh(\der{D})]$ the diagram on the right commutes. We
    will say that $\der{D}$are $\der{D}'$ are \emph{abstraction
      equivalent}, if there exists an abstraction equivalence between
    them.
  \end{definition}
}
\parbox{4cm}{\vspace{-1.35em}$\xymatrix@C=40pt{G'_i&D'_i \ar[r]^{g'_i}
    \ar@{>->}[l]_{f'_i}&G'_{i+1}\\ L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i
    \ar[u]^{k'_i} \ar[d]_{k_i} \ar[r]^{r_i} \ar@{>->}[l]_{l_i}
    &R\ar[u]^{h'_i} \ar[d]_{h_i}\\G_i
    \ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole&D_i\ar@{>->}[l]^{f_i}\ar@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole
    \ar[r]_{g_i}&G_{i+1}\ar@/_.45cm/[uu]_{\phi_{H_i}}}$}

@@

\begin{restatable}{proposition}{propUnique}
  \label{prop:unique}
  Let $\X$ be a $\mathcal{M}$-adhesive category. Suppose that the two
  direct derivations $\dder{D}$ and $\dder{D}'$ below, with the same
  match and applying the same left-linear rule $\rho$ are given.
  \[\xymatrix{L \ar[d]_{m}& K \ar[d]^{k}\ar@{>->}[l]_{l} \ar[r]^{r} &
      R \ar[d]^{h} & L \ar[d]_{m}& K \ar[d]^{k'}\ar@{>->}[l]_{l}
      \ar[r]^{r} & R \ar[d]^{h'}\\G & \ar@{>->}[l]^{f} D \ar[r]_{g}& H
      & G & \ar@{>->}[l]^{f'} D' \ar[r]_{g'}& H'}\]
  Then there an abstraction equivalence between $\dder{D}$ and
  $\dder{D}'$, whose first component is $\id{G}$.
\end{restatable}

[Proof \ref{propUnique-proof}]

\iffalse
\subsection{Consistent permutations}

@@

    Let $(C, \{c_X\}_{X\in \Delta(\der{D})})$ be a cocone, since $\Delta(\der{D}')$ is a subdiagram of $\Delta(\der{D})$, we get another cocone $(c, \{c_X\}_{X\in \Delta(\der{D}')})$ which induces an arrow $c'\colon \lpro \der{D}' \rpro \to C$ such that
    \begin{align*}
      c'\circ \iota_{G_n} \circ f_n & =c_{G_n} \circ f_n \\&= c_{D_n}\\&= c_{H}\circ g_n
    \end{align*}
    Therefore the arrows $c'$ and $c_H$ induce a morphism $c\colon P\to C$ and the thesis now follows at once.

@@

    Let us now show that the square above is a pushout. Take two arrows $a\colon \tproi{D}{1}\to C$, $b\colon \tproi{D}{2}\to C$ such that
    \[a\circ \iota_{1, H}=b\circ \iota_{2, H}\]
    We can use the previous equality to define a cocone $(C, \{c_X\}_{X\in \Delta(\der{D})})$ putting:
    \[c_X:=\begin{cases}
        a\circ \iota_{1, X} & X\in \Delta(\der{D}_1) \\
        b\circ \iota_{2, X} & X\in \Delta(\der{D}_2)
      \end{cases}\]
    From this, we can deduce at once the existence of a unique $c\colon \tpro{D}\to C$ such that
    \[c\circ \iota_X = c_X\]
    By construction, for every $X\in  \Delta(\der{D}_1)$ and $Y\in  \Delta(\der{D}_2)$ we have
    \[\begin{split}
        c\circ p_1 \circ \iota_{1,X}&=c\circ \iota_{X}\\&=c_X \\&=a\circ \iota_{1,X}\\&=a\circ p_1\circ \iota_{1,X}
      \end{split}\qquad \begin{split}
        c\circ p_2 \circ \iota_{2,Y}&=c\circ \iota_{Y}\\&=c_Y \\&=b\circ \iota_{2,Y}\\&=b\circ p_2\circ \iota_{2,Y}
      \end{split}\]
    Therefore
    \[c\circ p_1=a \qquad c\circ p_2 = b\]

    For uniqueness, suppose that $c'\colon \tpro{D}\to C$ is such that
    \[c'\circ p_1=a \qquad c'\circ p_2 = b\]
    Then, for every $X\in \Delta(\der{D})$ we have
    \begin{align*}
      c'\circ \iota_X & = \begin{cases}
        c'\circ p_1\circ \iota_{1, X} & X\in \Delta(\der{D}_1) \\
        c'\circ p_2\circ \iota_{2, X} & X\in \Delta(\der{D}_2)
      \end{cases} \\&=\begin{cases}
        a\circ \iota_{1, X} & X\in \Delta(\der{D}_1) \\
        b\circ \iota_{2, X} & X\in \Delta(\der{D}_2)
      \end{cases}\\&=c_X\\&=c\circ \iota_X
    \end{align*}
    showing that $c'=c$ as wanted.	 \qedhere
  \end{enumerate}
\end{proof}

\begin{corollary}
  \label{cor:colim}
  Let $\der{D}=\{\dder{D}_i\}_{i=0}^n$ a derivation of length $n+1$
  and fix an index $j\in[0,n]$. Define
  \[\der{D}^j_1:=\{\dder{D}_i\}_{i=0}^{j-1} \qquad
    \der{D}^j_2=\{\dder{D}_j\} \qquad
    \der{D}^j_3:=\{\dder{D}_i\}_{i=j+1}^n
  \]
  with the convention that
  $\der{D}^0_1$ and $\der{D}^n_3$ are the empty derivation on,
  respectively, $G_0$ and $G_n$. Then the square below is a pushout
  and a pullback
  \[\xymatrix@C=30pt{D_{j} \ar[ddrr]^{\iota_{D_j}}
      \ar[r]^{g_j}\ar@{>->}[d]_{f_j}& G_{j+1}
      \ar[ddr]^{\iota_{G_{j+1}}}
      \ar[r]^-{\iota_{3, G_{j+1}}} & \lpro \der{D}^j_3\rpro \ar[dd]^{p_2} \\
      G_j\ar[drr]_{\iota_{G_{j}}} \ar@{>->}[d]_{\iota_{1,G_j}}\\ \lpro
      \der{D}^j_1 \rpro \ar[rr]_{p_1} &&\tpro{D} }
  \]
  Where the two arrows $p_1\colon \lpro\der{D}^j_1 \rpro\to \tpro{D}$,
  $p_2\colon \lpro \der{D}^j_3\rpro \to \tpro{D}$ are induced by the
  cocones $(\tpro{D}, \{\iota_{X}\}_{X\in \Delta(\der{D}^j_1)})$ and
  $(\tpro{D},\{\iota_{X}\}_{X\in \Delta(\der{D}^j_3)})$, respectively.
\end{corollary}
\begin{remark}
  If $\der{D}$ is empty then $\der{D}^j_1, \der{D}^j_2$ and
  $\der{D}^j_3$ are empty too.
\end{remark}

\begin{proof}
  We can notice that
  $\der{D}=\der{D}^j_1\cdot \der{D}^j_2 \cdot \der{D}^j_3$. By the
  first and the second point of \Cref{lem:colim} then we get the
  following diagram, in which all squares are $\mathcal{M}$-pushouts.

  \[
    \xymatrix@C=30pt{D_{j} \ar[r]^{g_j}\ar@{>->}[d]_{f_j}& G_{j+1}
      \ar[r]^-{\iota_{3,G_{j+1}}} \ar[d]_{\iota_{2, G_{j+1}}}
      \ar@/^.5cm/[dd]^{\iota_{1,2, G_{j+1}}}& \lpro \der{D}^j_3\rpro
      \ar[dd]^{p_2} \\ G_j \ar[r]^{\iota_{2,
          G_j}}\ar@{>->}[d]_{\iota_{1,G_j}} & \lpro \der{D}^j_2\rpro
      \ar[d]_a\\ \lpro \der{D}^j_1 \rpro \ar@/_.4cm/[rr]_{p_1}
      \ar[r]^b &\lpro \der{D}^j_1\cdot \der{D}^j_2 \rpro
      \ar[r]^c&\tpro{D} }
  \]
  Applying \Cref{lem:po1} twice we get that the whole square is an
  $\mathcal{M}$-pushout. Then the thesis follows from
  \Cref{prop:pbpoad}.
\end{proof}

\begin{remark}
  \label{rem:zero1}
  In particular, considering $j=0$ or $j=n$, we get that the following
  two squares are $\mathcal{M}$-pushouts and, thus, pullbacks.
  \[
    \xymatrix{D_0 \ar[r]^-{\iota_{3, D_0}} \ar@{>->}[d]_{f_0}& \lpro
      \der{D}^0_3 \rpro \ar@{>->}[d]^{p_2} &D_n
      \ar@{>->}[d]_{\iota_{1, D_n}}\ar[r]^{g_n}& G_n
      \ar@{>->}[d]^{\iota_{G_n}}\\ G \ar[r]_{\iota_G}& \tpro{D} &
      \lpro \der{D}^n_1 \rpro \ar[r]_{p_1} & \tpro{D}}
  \]
\end{remark}

\begin{corollary}
  \label{cor:ele}
  Let $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ be a derivation between $G$
  and $H$. Let $j$ and $k$ be two indexes less or equal than $n+1$ and
  suppose that $j< k$.  Consider two arrows $a\colon T\to G_j$,
  $b\colon T\to G_k$. If $\iota_{G_j}\circ a = \iota_{G_k}\circ b$,
  then there exist a unique arrow $c\colon T\to D_j $ such
  that \[f_j\circ c = a\qquad \iota_{D_j}\circ c =\iota_{G_k}\circ b\]
\end{corollary}
